---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# shiva

<!-- badges: start -->
[![R-CMD-check](https://github.com/asshah4/shiva/workflows/R-CMD-check/badge.svg)](https://github.com/asshah4/shiva/actions)
[![Codecov test coverage](https://codecov.io/gh/asshah4/shiva/branch/main/graph/badge.svg)](https://codecov.io/gh/asshah4/shiva?branch=main)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

When looking at cardiac electrogram data, an important practice is identifying intervals between depolarizations. These range from _sinoatrial_ conduction properties, _atrioventricular_ and _ventriculoatrial_ intervals, _Hissian_ intervals, etc. These names (STIMULATION, HISSIAN, INTERVALS, VENTRICULAR, ATRIAL) were put into an anagram for the name of the package: `shiva`. 

The goal of `shiva` is to work with electrophysiology (EP) signal data to help evaluate intervals, pacing maneuvers, stimulation protocols, as well as generate informative plots for teaching/learning and understanding of EP studies.

## Installation

You can install the released version of shiva from [CRAN](https://CRAN.R-project.org) with:

```{r, eval=FALSE}
#install.packages("shiva")
```

And the development version from [GitHub](https://github.com/) with:

```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("asshah4/shiva")
```
