---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# EGM

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/EGM)](https://CRAN.R-project.org/package=EGM)
[![R-CMD-check](https://github.com/shah-in-boots/EGM/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/shah-in-boots/EGM/actions/workflows/R-CMD-check.yaml)
[![test-coverage](https://github.com/shah-in-boots/EGM/actions/workflows/test-coverage.yaml/badge.svg)](https://github.com/shah-in-boots/EGM/actions/workflows/test-coverage.yaml)
[![pkgdown](https://github.com/shah-in-boots/EGM/actions/workflows/pkgdown.yaml/badge.svg)](https://github.com/shah-in-boots/EGM/actions/workflows/pkgdown.yaml)
<!-- badges: end -->

When looking at cardiac electrogram data, an important practice is identifying intervals between depolarizations. These range from _sinoatrial_ conduction properties, _atrioventricular_ and _ventriculoatrial_ intervals, _Hissian_ intervals, etc. 

The goal of `EGM` is to work with electrophysiology (EP) signal data to help evaluate intervals, pacing maneuvers, stimulation protocols, as well as generate informative plots for teaching/learning and understanding of EP studies.

## Installation

You can install the released version of EGM from [CRAN](https://CRAN.R-project.org) with:

```{r, eval=FALSE}
#install.packages("EGM")
```

And the development version from [GitHub](https://github.com/) with:

```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("shah-in-boots/EGM")
```
