---
title: "Annotator Development and the WFDB-EGM Profile"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
vignette: >
  %\VignetteIndexEntry{Annotator Development and the WFDB-EGM Profile}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
library(EGM)
```

# Status and scope

This vignette proposes **WFDB-EGM version 0.1.0**, a draft annotation profile for
intracardiac electrograms. It is intended to become an implementation contract,
an interoperable research format, and a starting point for clinical consensus.
It is not yet a ratified clinical standard.

The proposal keeps the ordinary WFDB annotation file format. It adds a
versioned, namespaced vocabulary for information that the surface-ECG label set
cannot express: local activation methods, intracardiac components, near- and
far-field classification, morphology, bounded phenomena, anatomy, signal
quality, mechanical events, and multichannel relationships.

There is an important implementation distinction in the current package:

- `annotation_table()` and native WFDB annotation input/output are implemented.
- An exported `annotator` class, registry, profile validator, and EGM
  encoder/decoder are **not yet implemented**. Earlier drafts of this vignette
  described that class as if it already existed.

The code examples that only construct an `annotation_table` work today. APIs
marked as proposed define the work needed before the profile can be registered
and enforced by `{EGM}`.

## Design goals

WFDB-EGM should:

1. remain readable and writable as a WFDB-compatible MIT-format annotation
   file;
2. preserve the exact sample, channel, and acquisition context of an
   observation;
3. work for unipolar, bipolar, multipolar, and surface channels;
4. represent points, ranges, state changes, and multichannel events;
5. separate what was observed from what was inferred;
6. retain the method, reference, and confidence behind an annotation;
7. support new EP concepts without redefining existing WFDB labels; and
8. round-trip without access to a proprietary mapping system.

The first version does not attempt to define a new waveform file format, a
complete electroanatomic mapping format, or a fixed set of ablation targets.
Coordinates, catheter geometry, and large provenance objects may be linked as
external metadata, but the core annotation remains useful without them.

# What WFDB provides

In WFDB terminology an *annotator* is the suffix identifying one annotation set:
`record.egm` has the annotator name `egm`. It is not the same thing as the
proposed R `annotator` class, which will describe how a profile is validated,
encoded, decoded, and displayed. The name `atr` is reserved for reference
annotations supplied by a database's creator.

The MIT annotation format stores point annotations in sample order. The WFDB
API exposes `time`, annotation type, `subtyp`, `chan`, `num`, and `aux`.
The official [annotation structure documentation](https://physionet.org/physiotools/wpg/wpg_30.htm),
[binary format](https://physionet.org/physiotools/wag/annot-5.htm), and
[annotation-order rules](https://physionet.org/physiotools/wpg/wpg_48.htm)
impose the following constraints.

| `{EGM}` column | Wire meaning | Portable constraint | WFDB-EGM rule |
|---|---|---:|---|
| `sample` | Sample from record start | Integer | Canonical time; `time` is display-only |
| `type` | Annotation code | One supported WFDB type | A carrier hint, not the complete EGM meaning |
| `subtype` | Per-annotation modifier | Signed byte, -128 to 127 | Follow carrier semantics; WFDB-EGM reserves NOTE subtype 1, otherwise use 0 |
| `channel` | Annotation channel | Unsigned byte, 0 to 255 | Zero-based WFDB signal index for channel-scoped rows |
| `number` | Annotation number | Signed byte, -128 to 127 | Disambiguates annotations at the same sample and channel |
| `aux` | Auxiliary ASCII payload | At most 255 bytes for portable WFDB APIs | Versioned `EGM1|...` payload |

Although WFDB defines `number` as signed, writers targeting both C WFDB and
`wfdb-python` should prefer 0 through 127 unless negative-value interoperability
has been verified. The examples use only nonnegative disambiguators.

WFDB has no native duration or event identifier. A range therefore requires two
annotations, and a multichannel event requires multiple rows. The tuple
`(sample, number, channel)` identifies a WFDB annotation location. Files should
be sorted by `sample`, then `number`, then `channel`; two annotations at the
same location are not portable because canonical sorting may retain only the
last one.

Full WFDB-EGM conformance requires C WFDB 10.4.12 or later, which introduced
canonical `time`, `num`, then `chan` ordering. Older readers can preserve many
rows, but the profile must not rely on same-sample, same-channel rows being
distinguished by `number` when targeting those versions.

WFDB annotation files may declare an annotation time resolution distinct from
the record frame frequency. WFDB-EGM 0.1.0 requires them to be equal so `sample`
always means a record sample. A reader must recognize the standard
`## time resolution:` preamble and reject a mismatch until explicit rescaling
is implemented; the current native `{EGM}` reader does not yet do this.

Although the binary length field can represent a larger value, portable WFDB C
interfaces expose an AUX length of at most 255 bytes and recommend ASCII. The
current `{EGM}` writer instead permits up to 1023 bytes and silently truncates a
longer value. A conforming WFDB-EGM validator must reject a payload longer than
255 bytes rather than depend on that behavior. AHA-format annotation conversion
is out of scope because it can reduce auxiliary content to six characters.

## Existing and custom annotation codes

WFDB reserves numeric codes 42 through 49 for user-defined annotation types and
supports self-describing *modification labels*. This is the official route for
custom mnemonics; see the [WFDB extension guidance](https://physionet.org/physiotools/wpg/wpg_70.htm)
and [`setannstr`/`setanndesc` documentation](https://physionet.org/physiotools/wpg/strtim.htm).

The current `{EGM}` native codec cannot yet use that route: it writes only the
hard-coded standard one-character symbols and reads codes 42 through 49 as
`?`. WFDB-EGM 0.1.0 therefore uses standard codes as **transport carriers** and
places the authoritative EGM meaning in AUX. Codes 42 through 49 should not be
allocated until `{EGM}` can read, write, and preserve modification labels. A
future compact encoding may add them without changing the clinical vocabulary.

| Situation | Carrier `type` | Rule |
|---|---:|---|
| EGM point, component, beat, mechanical event, or interpretation | `"` (`NOTE`) | Default carrier; an `EGM1` NOTE uses `subtype=1` |
| Waveform/component range | `(` / `)` (`WFON` / `WFOFF`) | Pair with the same `id` and `edge=start/end` |
| Signal-quality range | `(` / `)` (`WFON` / `WFOFF`) | Pair per channel or use `scope=record`; AUX carries quality semantics |
| Legacy signal-quality change | `~` (`NOISE`) | Use only with the standard subtype noisy-signal bit mask |
| Rhythm change | `+` (`RHYTHM`) | Preserve conventional WFDB rhythm text when legacy rhythm tools are required |
| Numeric measurement | `=` (`MEASURE`) | Follow the legacy annotator's documented string convention or add a structured companion NOTE |
| External metadata | `@` (`LINK`) | AUX contains a URL understood by standard WFDB tools |

The traditional RHYTHM AUX value begins with `(`, for example `(VT` or `(AFIB`.
MEASURE AUX is an annotator-defined ASCII description; WFDB does not impose one
universal numeric/unit grammar. When legacy behavior and structured metadata
are both required, write the conventional RHYTHM or MEASURE row plus a
NOTE row at the same sample with a different `number`; link the NOTE through its
`id` or `group`. Do not concatenate `EGM1|...` onto a legacy value and assume
every WFDB application will parse it.

C WFDB treats leading sample-zero NOTE rows with `subtype=0` as its annotation
translation-table and time-resolution preamble, whether or not the AUX value is
a valid definition. It may therefore hide an ordinary NOTE at that location.
WFDB-EGM reserves NOTE `subtype=1` for every structured `EGM1` NOTE, including
the profile declaration, so those rows remain visible. Recognized standard
modification-label rows retain subtype 0 and are parsed as file metadata rather
than clinical annotations.

Existing symbols must not be reinterpreted. In particular, `a` means an
aberrated atrial premature beat, `V` means a premature ventricular contraction,
`f` means fusion of a paced and normal beat, `*` means systole, `D` means
diastole, and `!` means a ventricular flutter wave. The `|` artifact code
specifically means an isolated QRS-like artifact, not a generic catheter or
valve bump. None of these is a generic atrial, ventricular, fascicular,
diastolic-potential, fractionation, or mechanical-event label.

# A compositional clinical model

Intracardiac EGM concepts are not mutually exclusive. The same observed
component can be ventricular, near-field, late, isolated, fractionated, and a
local abnormal ventricular activity (LAVA). Flattening those properties into a
single `type` would lose information and create an unmaintainable number of
codes.

WFDB-EGM instead separates six axes:

| Axis | Question | Examples |
|---|---|---|
| Topology | What is its shape in time? | point, bounded component, interval, state |
| Observation | What was recorded? | activation, A/H/V component, QS, double potential, noise |
| Acquisition | How was it recorded or selected? | unipolar, bipolar, filter, max negative dV/dt, manual |
| Context | Where and during what? | channel, chamber, recording site, rhythm, reference window |
| Relation | What belongs together? | same beat, onset/offset pair, conducts to, evoked by |
| Interpretation | What might it mean? | near-field, focal origin candidate, block, critical isthmus |

## Observation before interpretation

An annotation should first preserve the waveform observation and measurement.
Interpretations are additional fields with evidence and confidence. For
example:

- two discrete deflections and their intervening gap are observations;
- `double` or `split` is a morphology classification;
- `line_of_block` is an interpretation that needs spatial or pacing evidence;
- a component recorded during VT diastole is an observation;
- `critical_isthmus` is an interpretation that requires more than timing alone.

This separation allows two reviewers to agree on component times while
disagreeing on mechanism. It also permits an algorithm's interpretation to be
revised without moving the underlying manual fiducials.

## Local activation time and QS morphology

For a unipolar EGM, the conventional local activation-time (LAT) fiducial is the
steepest negative intrinsic deflection. Bipolar timing is less uniform: onset,
first sharp component, peak, rapid downslope, and maximum absolute slope are all
used. Bipolar morphology and amplitude also depend on electrode order,
wavefront direction, electrode spacing, and filtering. These limitations are
summarized in the [EHRA/HRS ventricular-ablation consensus](https://academic.oup.com/europace/article/11/6/771/478564)
and the [EHRA position paper on EGM technology](https://academic.oup.com/europace/article/24/2/313/6456176).

The unipolar search must be restricted to the intended local component and
declared window; it is not simply the most negative slope anywhere in the
trace. In multicomponent VT substrate EGMs, a larger far-field ventricular
component can otherwise win. In a direct comparison, fixed-window unipolar
negative-dV/dt annotation selected only 33% of late potentials correctly
([Hawson et al.](https://doi.org/10.1016/j.jacep.2023.10.014)).

The fiducial vocabulary must therefore be explicit:

| `fiducial` value | Meaning |
|---|---|
| `uni_max_neg_dvdt` | Most negative unipolar slope in the declared search window |
| `bi_first_sharp` | First sharp bipolar component meeting the declared rule |
| `bi_peak` | Selected bipolar peak |
| `bi_max_abs_dvdt` | Bipolar sample with greatest absolute slope |
| `bi_max_neg_dvdt` | Most negative bipolar slope |
| `bi_max_pos_dvdt` | Most positive bipolar slope |
| `first_deflection` | First qualifying departure from baseline |
| `last_component` | Terminal qualifying component |
| `onset` / `offset` | Boundary rather than an activation peak |
| `other` | Method is named in `method` |

An unqualified `max_dvdt` value is invalid because it does not state polarity or
whether the absolute value was used. Every LAT should also identify its
reference and search window, either directly or through linked annotations.
`bi_max_neg_dvdt` and `bi_max_pos_dvdt` additionally require the channel's
electrode order/polarity because reversing a bipole reverses the slope sign.

A unipolar QS pattern is a separate morphology observation. It can support a
focal-origin hypothesis because activation may spread away from the electrode,
but it is not the LAT fiducial and does not prove the successful ablation site.
Visual QS assessment had poor specificity in one prospective outflow-tract
series ([Niu et al.](https://pubmed.ncbi.nlm.nih.gov/29570888/)). WFDB-EGM stores
`feature=qs` separately from `fiducial=uni_max_neg_dvdt` and, when appropriate,
stores `interpretation=origin_candidate` with evidence and confidence.

## Component identity and recording site

The component's physiological identity is independent of catheter name and
recording location. A His catheter can contain atrial, His, and ventricular
components. A coronary-sinus channel can contain near-field CS musculature and
far-field atrial or ventricular activity.

Recommended `component` values are:

- `atrial`, `ventricular`, and `his`;
- `right_bundle`, `left_bundle`, and `purkinje`;
- operational `p1` and `p2` labels used in fascicular VT;
- `accessory_pathway_candidate`, including a candidate Kent potential;
- `pv_muscle`, `svc_muscle`, and `cs_muscle`;
- `unknown` or an extension term.

Use `site` for the recording location and `source` for the inferred anatomical
origin. When a controlled anatomy is available, add `site_system` and
`site_code`. Free-text anatomy is allowed after percent encoding, but portable
terms should be promoted to a versioned registry rather than embedded in the
core standard.

Specialized conduction-system and pathway labels are electrophysiological
interpretations, not histological proof. P1 substrate remains debated, P2 is
often a presystolic Purkinje-like potential, and ordinary atrial or ventricular
components can mimic accessory- or slow-pathway potentials. Preserve the raw
component timing and morphology, then add the candidate label with evidence and
confidence.

## Near-field and far-field

The `field` values are `near`, `far`, `mixed`, and `indeterminate`. The label
applies to a component, not permanently to an entire channel. Bipolar recording
reduces but does not eliminate far-field activity, and unipolar recording may
contain substantial remote activity.

Useful supporting evidence includes simultaneous unipolar/bipolar comparison,
spatial restriction to neighboring electrodes, differential pacing,
extrastimulus-induced separation, disappearance during ablation, dissociation,
and entrance/exit block. A validated pulmonary-vein study, for example, used
ablation response, dissociation, and pacing to establish near- versus far-field
labels ([Klemm et al.](https://doi.org/10.1007/s10840-010-9508-2)). Store that
support in `evidence` and use `confidence` when the distinction is uncertain.

## Fractionation, double potentials, and continuous activity

Fractionation is a bounded morphology, not one point and not automatically an
ablation target. The range should preserve:

- onset and offset;
- component count and individual component times when available;
- duration, peak-to-peak voltage, and interdeflection or isoelectric gaps;
- rhythm, activation rate, and analysis epoch;
- the named threshold/rule; and
- acquisition settings or a link to channel metadata.

Published definitions vary materially. Historical ventricular criteria were
specific to electrode geometry and filtering. The original complex
fractionated atrial electrogram (CFAE) definition included at least two
deflections or continuous prolonged activity, or a mean cycle length no greater
than 120 ms over a 10-second epoch
([Nademanee et al.](https://pubmed.ncbi.nlm.nih.gov/15172410/)). Later studies
changed component counts, duration thresholds, and epochs. A CFAE label must
therefore carry `rhythm`, `method`, and the applied rule; it is not a synonym
for every fractionated signal.

A double potential consists of two discrete components separated by an
isoelectric or low-amplitude interval. A split potential is a related
operational classification whose gap threshold varies across studies. Store the
two component times and `gap_ms`; record conduction block, collision, or
layered activation only as separate interpretations.

## Ventricular substrate observations

The profile should support these overlapping terms without treating them as
synonyms:

| Feature | Operational meaning to store |
|---|---|
| `late` | Local ventricular component after a named QRS-offset reference |
| `isolated` | Component separated from the main ventricular EGM by an isoelectric/low-amplitude gap |
| `lava` | Sharp, high-frequency ventricular potential distinct from far-field V, during or after it |
| `diastolic` | Component during electrical diastole in VT; include cycle phase and QRS intervals |
| `mid_diastolic` | Diastolic component meeting a named phase rule |
| `presystolic` | Component preceding the following QRS |
| `p1` / `p2` | Operational fascicular-VT labels; P2 is often presystolic and Purkinje-like |
| `evoked` | Response temporally linked to a stimulus, without implying decrement |
| `decrementing` | Component whose latency increases under a named extrastimulus/pacing protocol |

Late-potential definitions and isolation thresholds differ across protocols.
LAVA is broader than a post-QRS late potential and may be buried within the
far-field ventricular EGM. The original LAVA study describes elimination of
these signals as a substrate endpoint
([Jaïs et al.](https://pubmed.ncbi.nlm.nih.gov/22492578/)); the
[2019 VA-ablation consensus](https://doi.org/10.1016/j.hrthm.2019.03.002)
emphasizes the heterogeneity of late, split, and fractionated strategies.

Diastolic timing alone does not establish participation in a VT isthmus because
bystander tissue can also activate in diastole. Circuit roles such as entrance,
central isthmus, exit, and bystander should be stored in `interpretation` and
linked to evidence such as concealed entrainment, PPI minus TCL, stimulus-QRS
interval, resetting, termination, or pace-map match.

A decrementing annotation should name the pacing site, coupling interval,
baseline component latency, post-extrastimulus latency, and the protocol's
delay criterion. `evoked` alone must not be treated as evidence of decrement.

## Atrial, junctional, and conduction-system observations

In addition to atrial, His, and ventricular components, useful EP-study
annotations include candidate accessory-pathway/Kent and slow-pathway
potentials, PV/SVC/CS-musculature components, bundle and Purkinje potentials,
conduction direction, capture, block, dissociation, and decremental behavior.

AH, HV, HA, VA, AA, HH, VV, stimulus-to-component, PPI, and tachycardia cycle
length are measurements between explicit fiducials. They should reference the
two endpoint IDs rather than rely on proximity in the file. This is especially
important for fused or fractionated components.

For atrial tachycardia, preserve the earliest local atrial time relative to a
named P-wave or intracardiac reference, unipolar QS as supporting morphology,
fractionated duration and percentage of tachycardia cycle length, and the
fraction of the cycle covered by an activation map. These observations can
support focal, localized-reentrant, or macroreentrant interpretations but do not
define the mechanism by themselves. Diastolic and prolonged potentials may be
atrial or ventricular; `component` and `rhythm` provide that context. See the
[international atrial-tachycardia consensus](https://doi.org/10.1093/europace/euaf324).

## Voltage and substrate interpretation

Peak-to-peak bipolar or unipolar voltage is an observation. `low_voltage`,
`scar_candidate`, and `dense_scar` are rule-qualified classifications whose
thresholds vary with chamber, rhythm, electrode size and spacing, orientation,
contact, wall thickness, wavefront, and filtering. Store the measured voltage,
units, signal configuration, rhythm, and named threshold profile rather than a
bare scar label.

`unexcitable` is stronger than `dense_scar`: it requires a documented
noncapture protocol, including pacing location, output, pulse width, and the
observed response. A voltage map alone cannot establish inexcitability.

## Surface references, beats, and rhythms

WFDB-EGM complements rather than replaces the standard surface-ECG vocabulary.
Use existing WFDB beat and wave labels when their published meanings fit. A
surface P onset/peak, QRS onset/peak/offset, or T-wave marker can also receive an
EGM ID through a companion NOTE so intracardiac activations can name it as
`reference`.

Recommended rhythm values include `sinus`, `af`, `atrial_flutter`,
`atrial_tachycardia`, `avnrt`, `avrt`, `junctional_tachycardia`, `vt`, `vf`, and
`indeterminate`. Rhythm is a record or channel-set state, not a property that
must be copied onto every channel. Beat groups may be classified as sinus,
atrial, ventricular, junctional, premature, escape, paced, fusion, capture, or
noncapture while preserving their channel-local A/H/V sequence.

Do not use a bare `paced` context for mapping. Distinguish `atrial_paced`,
`rv_paced`, `lv_paced`, `biventricular_paced`, `his_paced`, and
`lbb_area_paced`, and store pacing site, capture type, and cycle length.
Changing the paced wavefront can change voltage, fractionation, LAVA separation,
and LAT.

## Stimuli, maneuvers, and therapy

Pacing and procedural events are essential context for EP interpretation. A
stimulus is a point with its pacing site/channel, output, pulse width, train or
extrastimulus position, and group ID when available. Its responses are separate
activation/component annotations linked with `evoked_by` or `captured_by`.

Pacing trains, entrainment, programmed stimulation, RF, cryo, pulsed-field
ablation, and cardioversion are `procedure` ranges or events. Their electrical
artifact, tissue response, capture, termination, and post-procedure signal
change remain separate observations. This prevents an RF interval, for example,
from being confused with the saturation artifact it produces.

Phrenic-nerve capture is a stimulus response. Link it to the pacing stimulus;
encode the resulting diaphragmatic activation, motion, or electromyogram as a
separate biological/mechanical observation or artifact rather than collapsing
both into generic skeletal-muscle noise.

## Noise, artifact, and mechanical events

Signal quality is normally a range and may be channel-specific or record-wide.
Recommended `cause` terms include:

- `powerline`, `baseline_drift`, `high_frequency_noise`, and `crosstalk`;
- `clipping`, `saturation`, `dropout`, and `electrode_fault`;
- `filter_ringing`, `pacing_recovery`, `rf_artifact`, `pfa_artifact`, and
  `defibrillation_artifact`;
- `catheter_motion`, `poor_contact`, `respiration`, `skeletal_muscle`, and
  `unknown`.

Store `state=limited` or `state=unusable` rather than assuming every marked
range should be discarded. Filtering, electrode geometry, contact, and
recording-system noise materially alter EGM morphology and can create apparent
fractionation; see the [HRS EP-laboratory standards](https://doi.org/10.1016/j.hrthm.2014.03.042)
and a contemporary [recording-system comparison](https://pmc.ncbi.nlm.nih.gov/articles/PMC10658352/).

`catheter_bump`, `valve_contact`, `wall_contact`, `catheter_movement`, and
`catheter_displacement` are extensible `mechanical` events. They do not yet have
a widely standardized waveform definition, so record whether each was observed,
reported, or inferred. A mechanical event, its artifact range, and any induced
ectopy or transient conduction block are distinct annotations joined by a
group ID.

# The `EGM1` AUX envelope

The authoritative profile payload is a compact ASCII envelope:

```text
EGM1|kind=activation&id=act17&component=ventricular&field=near&fiducial=bi_max_abs_dvdt
```

The encoding rules are:

1. The exact prefix is `EGM1|`.
2. The remainder is an ampersand-separated list of `key=value` pairs.
3. Keys use lowercase `snake_case`.
4. Text is converted to UTF-8 bytes before URI percent encoding. Characters
   outside letters, digits, `.`, `_`, `~`, and `-` are encoded, and percent
   escapes use uppercase hexadecimal. A comma separates controlled-list values;
   a literal comma in free text is `%2C`. The resulting AUX payload is ASCII.
5. The complete AUX value must contain no NUL or control characters and must be
   no more than 255 ASCII bytes.
6. `kind` and `id` are required except where an exact legacy WFDB convention is
   intentionally used.
7. An ID uniquely names one logical entity within the annotation file. It
   contains at most 32 characters from letters, digits, `.`, `_`, `~`, and `-`.
   The same ID may repeat on paired boundaries or other rows that deliberately
   encode that entity; two different entities cannot share it.
8. Readers ignore but preserve unknown keys. A breaking grammar change requires
   a new prefix such as `EGM2|`.
9. Long notes, coordinate arrays, and provenance documents belong in linked
   metadata rather than AUX.
10. Duplicate keys and empty keys are invalid. Controlled clinical vocabulary
    values use lowercase; case-sensitive unit symbols and external code-system
    identifiers are exempt. Decimal numbers use `.` and carry an explicit unit
    key when the unit is not fixed by the field name.

Common keys are:

| Key | Meaning |
|---|---|
| `kind` | `profile`, `activation`, `component`, `measurement`, `beat`, `rhythm`, `stimulus`, `procedure`, `quality`, `mechanical`, `relation`, `anatomy`, or `detail` |
| `id` | Identity of one logical point, component, or range |
| `group` | Parent beat, maneuver, episode, or multichannel event |
| `parent` | Primary entity ID for a detail row or child observation |
| `part` / `parts` | One-based detail-row position and total detail-row count |
| `scope` | Omitted for a channel row; `record` when the carrier channel must be ignored |
| `edge` | `start` or `end` for a paired range |
| `component` | Physiological component identity |
| `feature` | Comma-separated observed morphologies |
| `field` | `near`, `far`, `mixed`, or `indeterminate` |
| `fiducial` | Exact LAT/boundary rule |
| `class` / `state` / `cause` | Beat class, state value, or quality/mechanical cause |
| `direction` | `antegrade`, `retrograde`, `bidirectional`, or `indeterminate` |
| `reference` | ID of the timing reference |
| `from` / `to` | Endpoint IDs for a measurement or relation |
| `site` / `source` | Recording location and inferred origin, respectively |
| `metadata_ref` | ID of a standard LINK row containing bulky external metadata |
| `rhythm` | Rhythm during the observation |
| `method` | Manual rule or algorithm identifier and version |
| `rule` | Named operational definition or threshold profile |
| `provenance` | `observed`, `reported`, `inferred`, or an algorithm identifier |
| `evidence` | Comma-separated support for an interpretation |
| `interpretation` | Revisable clinical inference |
| `confidence` | Number from 0 through 1 |

Measurements use explicit keys such as `slope`, `slope_unit`, `voltage_pp`,
`voltage_unit`, `duration_samples`, `duration_ms`, `gap_ms`, `cycle_ms`,
`window_start_ms`, `window_end_ms`, `n_deflections`, `value`, and `unit`. Units
are part of the value's meaning; never put an unlabelled number in AUX,
`subtype`, or `number`.
Unit values are case-sensitive; for example, millivolts are `mV`, not `mv`.

All conformance-required keys for the primary entity must fit in its first AUX
payload. Optional metadata that does not fit uses NOTE `subtype=1` detail rows
at the same sample and channel with distinct `number` values. Each detail row
has its own `id`, `kind=detail`, `parent=<primary_id>`, and `part`/`parts`; a
key-value pair must never be split across rows. Bulky provenance, coordinate
arrays, or free text use a standard LINK row and a `metadata_ref` key instead
of chained AUX fragments. Every individual row remains within 255 bytes.

## Profile declaration

Every WFDB-EGM annotation file, whatever its annotator suffix, should contain a
sample-zero profile NOTE. A typical payload is:

```text
EGM1|kind=profile&id=profile&scope=record&profile=wfdb_egm&version=0.1.0&channel_base=0
```

The profile row uses NOTE `subtype=1` and the first available nonnegative
`number` at sample 0, channel 0 so it survives C WFDB preamble parsing and does
not collide with a real annotation at that location.

The annotator suffix and the profile are separate. Use `egm` for one canonical
WFDB-EGM annotation set. When a record contains multiple independent sets, use
portable alphabetic creator names such as `egmmanual` and `egmalgo`, and declare
the same profile in each file. Alphabetic suffixes avoid current
[`wfdb-python` restrictions](https://github.com/MIT-LCP/wfdb-python/blob/main/wfdb/io/annotation.py)
on digits in annotation extensions. Do not use `atr` unless the
file is the database creator's reference set.

A [PhysioNet-style collection](https://archive.physionet.org/tutorials/creating-records.shtml)
lists each suffix and a short description in its `ANNOTATORS` file, for example:

```text
egm	WFDB-EGM 0.1.0 intracardiac electrogram annotations
```

## Channel convention and global scope

For a channel-scoped annotation, WFDB-EGM defines `channel` as the **zero-based
WFDB signal index**. Anatomy does not belong in this byte. Static catheter,
electrode, polarity, and channel information belongs in the WFDB header or
linked metadata; a catheter-position change is a time-varying annotation.

A record-wide row uses `scope=record`; its carrier `channel` is ignored and
should be 0. Consequently, a reader must use `scope`, not `channel == 0`, to
identify a global annotation.

This exposes a current package incompatibility that must be resolved before
conformance: `header_table$number` is one-based and existing helper
documentation treats annotation channel 0 as global. The WFDB-EGM encoder must
subtract one when mapping a physical `header_table` channel to the wire, and
`add_annotation()` must stop inferring global scope from channel 0 alone.

## A minimal conforming table

A file-level table starts with its profile row and includes every referenced
entity. This small example has a record-wide QRS-onset reference and one local
activation on the sixth WFDB signal.

```{r}
minimal_egm <- annotation_table(
  annotator = "egm",
  sample = c(0L, 125400L, 125430L),
  type = rep("\"", 3L),
  subtype = rep(1L, 3L),
  channel = c(0L, 0L, 5L),
  number = c(0L, 0L, 0L),
  aux = c(
    paste0(
      "EGM1|kind=profile&id=profile&scope=record",
      "&profile=wfdb_egm&version=0.1.0&channel_base=0"
    ),
    paste0(
      "EGM1|kind=activation&id=qrs42&scope=record",
      "&component=ventricular&feature=qrs",
      "&fiducial=onset&method=manual"
    ),
    paste0(
      "EGM1|kind=activation&id=act17&component=ventricular",
      "&field=near&fiducial=bi_max_abs_dvdt&reference=qrs42",
      "&window_start_ms=-100&window_end_ms=100&method=manual"
    )
  )
)

minimal_egm
```

With a matching `record.hea` present, the ordinary package I/O remains the
transport mechanism:

```{r}
#| eval: false
write_annotation(
  minimal_egm,
  annotator = "egm",
  record = "record",
  record_dir = "path/to/record"
)

roundtrip <- read_annotation(
  record = "record",
  annotator = "egm",
  record_dir = "path/to/record"
)
```

# Points and ranges

The following tables isolate individual patterns for readability. They are
fragments: a file written to disk also needs the profile row and any referenced
IDs or parent groups illustrated in the minimal conforming example above.

## A bipolar ventricular LAT

This table can be written by the current codec because it uses the standard
NOTE carrier. Channel 5 means the sixth WFDB signal.

```{r}
bipolar_lat <- annotation_table(
  annotator = "egm",
  sample = 125430L,
  type = "\"",
  subtype = 1L,
  channel = 5L,
  number = 0L,
  aux = paste0(
    "EGM1|kind=activation&id=act17&group=beat42",
    "&component=ventricular&field=near",
    "&fiducial=bi_max_abs_dvdt&slope=2.4",
    "&slope_unit=mV%2Fms&reference=qrs42",
    "&window_start_ms=-80&window_end_ms=20",
    "&method=manual&confidence=0.95"
  )
)

bipolar_lat
```

If a simultaneous unipolar channel has a QS pattern, record it independently:

```{r}
unipolar_qs <- annotation_table(
  annotator = "egm",
  sample = 125429L,
  type = "\"",
  subtype = 1L,
  channel = 6L,
  number = 0L,
  aux = paste0(
    "EGM1|kind=component&id=cmp18&group=beat42",
    "&component=ventricular&feature=qs",
    "&fiducial=uni_max_neg_dvdt",
    "&interpretation=origin_candidate",
    "&evidence=spatial_activation_map&confidence=0.70"
  )
)

unipolar_qs
```

The one-sample difference is retained rather than forcing the two methods to
agree. Here `feature=qs` and `fiducial=uni_max_neg_dvdt` remain independent
properties even though they share one row. If the QS waveform extent matters,
encode it as a bounded component as well.

## A fractionation range

All bounded intervals are half-open: `[start, end)`. The two boundaries share
an `id`, channel, and group. Ranges may overlap or nest. A completed file must
close every start boundary.

```{r}
fractionation <- annotation_table(
  annotator = "egm",
  sample = c(210040L, 210097L),
  type = c("(", ")"),
  subtype = c(0L, 0L),
  channel = c(12L, 12L),
  number = c(0L, 0L),
  aux = c(
    paste0(
      "EGM1|kind=component&id=frac7&group=beat88&edge=start",
      "&component=ventricular&feature=fractionated",
      "&rhythm=sinus&method=manual",
      "&rule=deflection_count_ge_4&n_deflections=5"
    ),
    paste0(
      "EGM1|kind=component&id=frac7&group=beat88&edge=end",
      "&component=ventricular&duration_samples=57"
    )
  )
)

fractionation
```

Individual deflections can be NOTE rows with their own IDs and
`group=frac7`. The LAT selection, if one is made, is another linked point rather
than an implicit property of the range.

## A channel-specific noise range

For many-channel recordings the historical WFDB NOISE subtype mask is too
small. Use paired WFON/WFOFF boundaries per affected channel and join them with
an ID. Reserve `~` for its standard bit-mask meaning.

```{r}
noise_range <- annotation_table(
  annotator = "egm",
  sample = c(310000L, 312500L),
  type = c("(", ")"),
  subtype = c(0L, 0L),
  channel = c(18L, 18L),
  number = c(0L, 0L),
  aux = c(
    paste0(
      "EGM1|kind=quality&id=noise3&edge=start",
      "&state=unusable&cause=catheter_motion&method=manual"
    ),
    "EGM1|kind=quality&id=noise3&edge=end&state=usable"
  )
)

noise_range
```

# Multichannel beats and events

One physiological event does not imply one identical timestamp on every
channel. A junctional beat can have a His component, ventricular activation,
and retrograde atrial activation at different samples. Preserve those
differences and connect the rows with `group`.

```text
beat42
|-- His channel: H at t0
|-- RV channel: V at t0 + HV
|-- HRA channel: retrograde A at t0 + HA
|-- CS channels: A components at their local activation times
`-- surface ECG: QRS reference
```

```{r}
junctional_beat <- annotation_table(
  annotator = "egm",
  sample = c(500000L, 500000L, 500042L, 500071L),
  type = rep("\"", 4L),
  subtype = rep(1L, 4L),
  channel = c(0L, 3L, 7L, 1L),
  number = c(0L, 0L, 0L, 0L),
  aux = c(
    paste0(
      "EGM1|kind=beat&id=beat42",
      "&scope=record&class=junctional&provenance=inferred",
      "&evidence=ep_maneuvers&confidence=0.95&method=manual"
    ),
    paste0(
      "EGM1|kind=activation&id=h42&group=beat42",
      "&component=his&fiducial=bi_first_sharp"
    ),
    paste0(
      "EGM1|kind=activation&id=v42&group=beat42",
      "&component=ventricular&fiducial=bi_first_sharp"
    ),
    paste0(
      "EGM1|kind=activation&id=a42&group=beat42",
      "&component=atrial&direction=retrograde",
      "&fiducial=bi_first_sharp"
    )
  )
)

junctional_beat
```

This example assumes the junctional diagnosis was adjudicated using EP
maneuvers not reproduced in the four rows. A/H/V order alone does not establish
junctional origin; without that evidence use `class=indeterminate` and retain
the observed component sequence.

At one sample and channel, multiple rows receive distinct `number` values. The
number is only a local disambiguator; it is not an event ID and should not be
used to link channels.

Useful relationship terms include `component_of`, `same_beat_as`, `precedes`,
`follows`, `conducts_to`, `blocked_to`, `dissociated_from`,
`far_field_projection_of`, `evoked_by`, `captured_by`, `terminates`, and
`causes_artifact`. A relation names its endpoint IDs in `from` and `to`.

# Acquisition and anatomical metadata

Reproducible interpretation needs more than a channel label. Record or link the
following static metadata for every channel when available:

- surface versus intracardiac modality;
- unipolar, bipolar, or derived configuration;
- positive and negative electrodes and unipolar reference;
- catheter and electrode identifiers, electrode size, and spacing;
- chamber and baseline catheter position;
- sampling frequency, gain, units, filter settings, and notch-filter state;
- whether the signal is raw or processed; and
- mapping-system or recording-system provenance.

WFDB signal descriptions, header information strings, and the existing gain and
unit fields should contain what they can. A richer record-level manifest may be
linked with the standard `LINK` annotation. Time-varying catheter location,
contact, or polarity changes remain annotations. Do not place patient
identifiers in AUX or in a public manifest.

# Proposed R `annotator` contract

An `annotation_table` should remain a seven-column R transport table: six wire
fields plus the derived display `time`. A profile decoder can expose a richer R
view, but its extra columns must be encoded back into AUX before
`write_annotation()` is called.

The proposed class interface is illustrative and is not yet exported:

```{r}
#| eval: false
wfdb_egm <- new_annotator(
  id = "wfdb_egm",
  version = "0.1.0",
  suffix = "egm",
  aux_prefix = "EGM1|",
  channel_base = 0L,
  note_subtype = 1L,
  carriers = c(
    note = "\"", onset = "(", offset = ")",
    legacy_noise = "~", rhythm = "+", measure = "=", link = "@"
  ),
  schema = egm_annotation_schema(),
  encode = encode_egm_annotation,
  decode = decode_egm_annotation,
  validate = validate_egm_annotation
)

register_annotator(wfdb_egm)
```

The class should provide:

1. `encode()` and `decode()` methods that are lossless for recognized and
   unknown AUX keys;
2. row validation for types, byte limits, units, IDs, and controlled terms;
3. file validation for profile declaration, canonical order, duplicate
   locations, paired boundaries, references, groups, and channel bounds;
4. an explicit conversion between one-based `header_table$number` and
   zero-based WFDB channels;
5. display metadata for points, spans, interpretations, and global states;
6. version negotiation and migration without changing old files; and
7. a registry that distinguishes the annotation-file suffix, profile ID,
   creator, and algorithm version.

The complete file must be decoded and its preamble, profiles, pairs, parents,
groups, and references validated **before** applying `begin`/`end` filtering.
Every interval intersecting a requested half-open window remains active in the
decoded view; preserve its original boundaries and mark a clipped/partial view
rather than orphaning one edge. Retain required out-of-window parent and
reference rows as context. A blindly row-filtered transport table is not a
conforming, rewritable window.

The current `read_annotation()` filters rows first, so it can discard the
sample-zero profile, one range edge, or an external reference. Profile-aware
windowing must occur above or replace that behavior.

Unknown fields must survive decode/encode. A strict reader may report an
unrecognized required term, but it must not erase it.

# Minimum conformance by use case

| Use case | Required representation |
|---|---|
| Local activation | Point, channel, component, explicit `fiducial`, reference/window, method |
| Unipolar QS | `feature=qs`; any origin claim is a separate interpretation |
| Near/far field | Component-level `field`, attributed source when known, evidence/confidence |
| Fractionation | Paired range, rhythm, named rule; duration/count/voltage when measured |
| Double/split potential | Component points, bounded parent, measured gap, named threshold |
| Late potential | Ventricular component and explicit QRS-offset reference/latency rule |
| VT diastolic potential | Component, VT cycle context, QRS intervals; circuit role optional |
| Noise/artifact | Paired quality range, channel or record scope, state and cause |
| Catheter/valve bump | Mechanical point/range and provenance; link any artifact or induced beat |
| Multichannel beat | Parent group plus channel-local component IDs and samples |
| Anatomy | Recording `site` separate from inferred `source`; code system when available |

# Implementation and validation roadmap

The profile should not be called implemented or interoperable until the
following work is complete.

## Transport hardening

- Document and enforce a scalar annotator/profile name for writing and profile
  validation, while retaining vector input only as a read convenience; enforce
  all column types.
- Preserve numeric custom annotation codes and modification labels, even before
  custom EGM codes are allocated.
- Reject AUX values over 255 bytes and embedded control characters.
- Sort and validate by `sample`, `number`, and `channel`, not sample alone.
- Parse standard sample-zero modification labels and annotation-clock metadata
  separately, then reject duplicate `(sample, number, channel)` locations among
  ordinary annotations.
- Require annotation resolution to equal record frame frequency in version 0.1.0,
  or implement explicit parsing and sample rescaling before relaxing that rule.
- Resolve channel-name conversion, zero-based wire indexing, and record scope.
- Require points and starts in `0..nsamp-1`; allow an end boundary at `nsamp` to
  close a half-open interval exactly at record end.

## Profile implementation

- Implement the `annotator` class, registry, `EGM1` parser, encoder, and rich
  decoded view.
- Implement pair, ID, group, relation, unit, confidence, and vocabulary
  validation.
- Add helpers for LAT, components, ranges, quality, mechanical events,
  measurements, and multichannel groups.
- Make plotting render focal points, channel spans, record-wide states, and
  revisable interpretations differently.

## Interoperability tests

A conformance fixture should pass all of these tests:

1. R write/read round trip of the six wire fields (`sample`, `type`, `subtype`,
   `channel`, `number`, and `aux`), with `time` checked after normalization from
   the record frequency.
2. C WFDB `rdann -f 0` inspection and `wrann` reconstruction.
3. Python `wfdb.rdann()` and `wfdb.wrann()` round trip.
4. A subtype-1 profile survives C preamble parsing and R/Python windowed reads,
   while standard subtype-0 modification labels remain metadata.
5. Same-time multichannel rows and same-channel multi-`number` rows.
6. AUX lengths of 0, 254, and 255 bytes, with rejection at 256 bytes.
7. Ranges beginning at sample 0, ending with an offset at `sample == nsamp`,
   overlapping, and nested.
8. Channel 0 as a physical signal, the highest physical channel, and a
   `scope=record` row.
9. Unknown AUX keys preserved exactly.
10. Rejection of AHA-format export as a lossless WFDB-EGM conversion.

## Clinical validation and governance

Before a 1.0 standard, a public, de-identified corpus should include atrial,
ventricular, His-Purkinje, accessory-pathway, pacing, ablation, noise, and
mechanical examples from multiple vendors and acquisition settings. At least
two EP reviewers should annotate difficult LAT, fractionation, and near/far
cases so agreement and adjudication can be measured.

Vocabulary changes need a public registry, semantic versioning, definitions,
examples, and a deprecation policy. Experimental values should use a namespaced
form such as `x_org_term`; widely adopted terms can later be promoted without
rewriting the original annotations.

# Evidence base

The profile is informed by the following standards and representative EP
literature:

- Moody, *WFDB Applications Guide*: [annotation format](https://physionet.org/physiotools/wag/annot-5.htm),
  [annotation structures](https://physionet.org/physiotools/wpg/wpg_30.htm),
  [annotation codes](https://physionet.org/physiotools/wpg/wpg_36.htm), and
  [extensions](https://physionet.org/physiotools/wpg/wpg_70.htm).
- Aliot et al., ventricular mapping and EGM interpretation consensus:
  [Europace 2009](https://academic.oup.com/europace/article/11/6/771/478564).
- Cronin et al., 2019 HRS/EHRA/APHRS/LAHRS ventricular-arrhythmia ablation
  consensus: [DOI 10.1016/j.hrthm.2019.03.002](https://doi.org/10.1016/j.hrthm.2019.03.002).
- van Schie et al., critical appraisal of unipolar/bipolar EGM technology and
  LAT during AF: [Europace 2022](https://academic.oup.com/europace/article/24/2/313/6456176).
- Hawson et al., comparison of VT LAT annotation methods:
  [DOI 10.1016/j.jacep.2023.10.014](https://doi.org/10.1016/j.jacep.2023.10.014).
- Niu et al., limitations and quantitative assessment of unipolar QS:
  [DOI 10.1111/jce.13491](https://doi.org/10.1111/jce.13491).
- Jaïs et al., LAVA definition and elimination:
  [DOI 10.1161/CIRCULATIONAHA.111.043216](https://doi.org/10.1161/CIRCULATIONAHA.111.043216).
- Bogun et al., isolated ventricular potentials and pace mapping:
  [DOI 10.1016/j.jacc.2005.12.062](https://doi.org/10.1016/j.jacc.2005.12.062).
- Saito et al., diastolic potentials and bystander sites in VT:
  [DOI 10.1023/A:1009776618809](https://doi.org/10.1023/A:1009776618809).
- Nogami et al., P1/P2 Purkinje potentials in fascicular VT:
  [DOI 10.1016/S0735-1097(00)00780-4](https://doi.org/10.1016/S0735-1097(00)00780-4).
- Nademanee et al., the original clinical CFAE definition:
  [DOI 10.1016/j.jacc.2003.12.054](https://doi.org/10.1016/j.jacc.2003.12.054).
- Klemm et al., validated pulmonary-vein near-/far-field discrimination:
  [DOI 10.1007/s10840-010-9508-2](https://doi.org/10.1007/s10840-010-9508-2).
- Haines et al., HRS EP-laboratory recording and signal-quality standards:
  [DOI 10.1016/j.hrthm.2014.03.042](https://doi.org/10.1016/j.hrthm.2014.03.042).
- Meyer et al., international atrial-tachycardia consensus:
  [DOI 10.1093/europace/euaf324](https://doi.org/10.1093/europace/euaf324).

These sources do not supply one universal definition for every EGM feature.
That variability is the reason WFDB-EGM records the rule, acquisition context,
evidence, and confidence rather than making a binary label carry more certainty
than the signal supports.
