---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(EGM)
```

# Reading in Data

Thie software supports reading in certain types of raw cardiac electric signal currently. 
This includes both intracardiac and extracardiac data.
Please see `read_muse()` and `read_bard()` as examples.
If additional formats are requested, please [file an issue](https://github.com/shah-in-boots/EGM/issues) and provide a sample file (which primarily requires signal, in whatever format, and meta-information about the individual channels).

Here is a simple example of taking a MUSE XML file and converting it to digital signal.
This was tested against the 9th version of the MUSE XML format.

```{r}
# Read in data
fp <- system.file('extdata', 'muse-sinus.xml', package = 'EGM')
xml <- readLines(fp)
head(xml)

# Instead, can read this in as a MUSE XML file
# Now as an `egm` class
ecg <- read_muse(fp)
ecg

# A quick look at one lead. Figures for presentation live in the companion
# {gram} package; {EGM} itself carries no plotting layer.
sig <- get_signal(ecg)
plot(sig$sample / 500, sig$II, type = "l", xlab = "Seconds", ylab = "Lead II")
```

Similarly, intracardiac recordings obtained through bard can be read in as well. The function itself documents how this generally work.

```{r}
# Read in data
fp <- system.file('extdata', 'bard-avnrt.txt', package = 'EGM')
bard <- readLines(fp)
head(bard, n = 20)

# Instead, read this as signal, breaking apart header and signal data
# Presented as an `egm` class object
egram <- read_bard(fp)
egram

# The intracardiac channels, one trace each, on a shared time axis
sig <- get_signal(egram)
matplot(
  sig$sample / 1000,
  sig[, c("HIS.D", "HIS.M", "CS.1.2", "RV.1.2")],
  type = "l", lty = 1,
  xlab = "Seconds", ylab = "Signal"
)
```

# Class Introduction

The `{EGM}` package introduces a specific data-oriented class called `egm` (note the *lower-case* spelling compared to the package name). 
This class is the primary data structure, and allows for compatibility within multiple signal types, including when reading in WFDB-formatted data.
Once in this class, everything can be written out as well - with the preferred option being for the stated `WFDB` format.
