---
title: "WFDB: An Introduction to the Waveform Database Software Package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{WFDB: An Introduction to the Waveform Database Software Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(EGM)
```

The package relies and is augmented heavily by the [Waveform Database (WFDB) software package](https://physionet.org/content/wfdb/10.7.0/), which is a well-documented and well-developed ecosystem of software applications, primarily written in `C` and `C++` that manage the storage, reading, writing, and interaction with electrical signal data.
This software is mostly external to the `{EGM}` package, and is used to supplement and expand the software.

__DISCLAIMER__: The software required to use annotations has not yet been re-written into a `C` backend for `R`, and thus requires using a local installation of `WFDB` for functionality. 
To build online vignettes and articles, where the software package is not available, I only demonstrate non-running examples.

# Installation

To install the `WFDB` software in its traditional, `C`-based format, I have found it easiest with teh instructions from the [Github](https://github.com/bemoody/wfdb) source. 
The installation instructions are relatively clear across multiple operating systems.

`WFDB` is easiest to install on a Unix-based system, such as Linux or MacOS. 
For Windows, I have found that using [WSL2](https://learn.microsoft.com/en-us/windows/wsl/install) is the most consistent and supported way to utilize the software.
When hidden on WSL, the path must be specified explicitly using the `set_wfdb_path()` function using the command from Powershell, for example, to switch to a WSL command environment.

```{r}
#| eval: false
set_wfdb_path("wsl /usr/local/bin")
```

Once this is in place, you should be able to work with `WFDB` files directly from `R` without significant overhead costs, while also gaining access to a variety of software applications.

# Annotations

The WFDB software package utilizes a binary format to store annotations.
Annotations are essentially markers or qualifiers of specific components of a signal, specifying both the specific time or position in the plot, and what channel the annotation refers to. 
Annotations are polymorphic, and multiple can be applied to a single signal dataset. 
The credit for this work goes directly to the original software creators, as this is just a wrapper to allow for flexible integration into `R`. 

To begin, let's take an example of a simple ECG dataset.
This data is included in the package, and can be accessed as below.

```{r}
fp <- system.file('extdata', 'muse-sinus.xml', package = 'egm')
ecg <- read_muse(fp)
fig <- ggm(ecg) + theme_egm_light()
fig
```

We may have customized software, manual approaches, or machine learning models that may label signal data.
We can use the `annotation_table()` function to create `WFDB`-compatible annotations, updating both the `egm` object and the written file.
To trial this, let's label the peaks of the QRS complex from this 12-lead ECG.

1. Create a quick, non-robust function for labeling QRS complex peaks
1. Evaluate the fit of the peaks to the dataset
1. Place the annotations into a table, updating the `egm` object
1. Plot the results

```{r}
# Let x = 10-second signal dataset
# We will apply this across the dataset
find_peaks <- function(x) {
}

# Create a signal dataset
dat <- extract_signal(ecg)
x <- dat$I
```
